### 2013-12-04 mcmc.tex
tuning section; some notes to self

% to-do
% -----
% - write zeroth draft
% - write first draft
% - send to friendlies for comments
%   - Lang
%   - Rix
%   - Bovy
% - post on arXiv

@@

...If you are really good, you will pass forward a sampling.  How big
should it be?  Refer to Mackay here.

@@

...What plots should you make?
Make plots that demonstrate a non-converged and a converged chain...

@@

...Return here to the proposal distribution, which exists for most methods...

...How do you make it run fast?..

...Acceptance ratio...

...ESJD and etc...

@@

...Gibbs sampling and interior marginalizations...

>>>>

% to-do
% -----
% - write zeroth draft
% - write first draft
% - send to friendlies for comments
%   - Lang
%   - Rix
%   - Bovy
%   - Brewer
% - post on arXiv

@@

...If you are really good, you will pass forward a sampling.
...How big should it be? Refer to Mackay here.
...Usefulness of a big sampling for hierarchical inferences; concept of interim prior?

@@

...What plots should you make?
...Plots showing individual walkers going through the space.
...(now show plots that demonstrate a non-converged and a converged chain)...

@@

@@

Most MCMC methods
  (though not all\note{We are promoters of an ensemble method that has a
  nearly-fixed proposal distribution and requires no tuning (CITE EMCEE).})
  make use of something like a ``proposal distribution''
  which determines what kinds of steps the walker can take
  as it random-walks through the parameter space.
The user generally has a lot of control over what this proposal distribution might be,
  and how to choose the parameters.
For example, in a $d$-dimensional parameter space, %do we consistently use d here?
  a zero-mean but anisotropic Gaussian (normal distribution)
  is often used to draw offsets for the walker to take from point to point.
Even within this choice (which is of one function among many possibilities),
  there are $d\,(d+1)/2$ parameters in the $d\times d$ symmetric, positive definite covariance matrix
  to set ``by hand''.
How to choose this function and set these parameters?

The key idea here is that if the proposal distribution is too narrow---%
  it proposes steps too small---%
  almost all steps will be accepted (REFERENCE ALGORITHM HERE),
  but it will take a long time to move anywhere because of timidity.
If the propsoal distribution is too wide---%
  it proposes steps too large---%
  the moves will cover parameter space easily,
  but almost no steps will be accepted;
  it will tend to jump to much lower probability regions.
There is a Goldilocks step size (proposal distribution root-variance) that is ``just right''.
In one dimension this might be easy to find,
  but, as we say, in large numbers of dimensions,
  there is a lot of freedom in choosing the parameters of the distribution.

In a deep sense, the only scalar that makes sense to optimize,
  when choosing proposal distribution (or, loosely speaking, step size),
  is the autocorrelation time, described above.
The optimal step size is the step size that makes for the shortest autocorrelation time.
This is easy to state, but hard to use in practice;
  being a second-order statistic of the MCMC chain,
  the autocorrelation time is a hard thing to measure without a lot of data,
  so it is hard to quickly estimate it and adjust,
  much less put it into some kind of optimization loop.
We usually use proxies for this of various kinds.

The simplest proxy for autocorrelation time optimization is the \emph{acceptance fraction}.
If you are accepting almost all proposed steps, your step sizes are too small on average.
If you are accepting almost none, your step sizes are too large.
The Goldilocks value is around a half,
  with a strong argument floating around that it should be 0.234 for best performance\note{CITE GELMAN}.
In the burn-in phase (discussed below) of an MCMC run,
  it makes sense to track the acceptance ratio,
  and adjust the proposal distribution variance as you get acceptance ratios
  that are far from the Goldilocks ratio.
This process can be automated easily;
  such automation is part of many projects that use MCMC.

A very common---%
  and very useful---%
  kind of proposal distribution is one that cycles through parameters,
  taking a random step in just one parameter at a time.
Aside from involving some persistence of state,
  this kind of proposal distribution can be valuable,
  in part because it reduces the (almost impossible) $d$-dimensional tuning problem
  to $d$ one-dimensional problems:
In this form of proposal distrubution,
  it is possible to track a separate acceptance fraction for every parameter.
Code can be built that uses the burn-in phase to tune all $d$ proposal variances
  such that each of them, individually,
  obtains acceptance at the same Goldilocks ratio.

One very important note to make here,
  because it is relevant to tuning,
  is that tuning can \emph{only} take place during the burn-in phase;
  you cannot tune \emph{while} you run your final MCMC run.
Why not?
Because tuning the proposal distribution based on the past history of the chain
  violates the ``Markov'' property
  that each step depends \emph{only} on the state at the previous step.
Violating the Markov property can be very bad;
  when you violate it,
  you lose all the provable properties of MCMC
  on which all our righteous power is based.

Another proxy for autocorrelation time is the
  Expected Squared Jump Distance.\note{(CITE GELMAN)}
This is the mean squared distance the walker moves, per step.
It is maximized when the acceptance ratio is reasonable
  and the step size is large;
  it is large when the exploration of the space is fast.
This proxy is easy to measure and use for tuning,
  and more directly related to autocorrelation time than the acceptance ratio.
We recommend using it for tuning, though we have never used it ourselves.
Like the acceptance fraction, it can also be used to tune
  in the case that the proposal distribution loops over parameters;
  once again, the user gets (in this case) $d$ one-dimensional tunings.

When sampling really gets very slow,
  there are various tricks and tips to work through.
One is to see if any of the parameters has a conditional posterior
  probability that is guaranteed to be Gaussian
  (this is true for some linear parameters in least-square problems
  when the priors are Gaussian, for example).
These dimensions can often be sampled \emph{exactly},
  outside of the MCCM framework;
  the algorithm looks like:
Take an MCMC step in the hard parameters,
  then take an exact step in the easy parameters,
  repeat.
Another related trick is to analytically marginalize,
  rather than sample,
  easy nuisance parameters.
The huge bag of possible tricks is so large,
  it goes way beyond the scope of this introductory \documentname.
In our experience, it is valuable to make friends with at least one statistician,
  one applied mathematician, and one computer scientist.
Between them, they ought to span the literature.
